---
title: "The Geographic Distribution of Urodynamics Testing Centers in the United States"
subtitle: "An Analysis of Medicare Data"
format:
  html:
    toc: true
    toc-expand: true
    toc-location: body
date: 2026-03-31
theme: lightly
author:
  - name: "Tyler Muffly, MD"
    affiliations: "Department of Obstetrics and Gynecology"
embed-resources: true
toc-title: Contents
page-layout: full
code-fold: true
editor: visual
editor_options:
  chunk_output_type: console
---

# Code Blocks

## Setup data

### Load libraries & functions

```{r}
packs <- c("readxl", "dplyr", "data.table", "janitor", "doParallel",
           "stringr", "Hmisc", "tidygeocoder", "tidycensus",
           "ggplot2", "ggpubr", "ggthemes", "ggbeeswarm",
           "geosphere", "sf", "usmap")
lapply(packs, require, character.only = TRUE)
```

### Import data

Medicare Provider Utilization and Payment Data filtered for urodynamics testing
CPT codes:

- 51726: Complex cystometrogram (CMG)
- 51727: CMG with voiding pressure studies
- 51728: Complex CMG with voiding pressure studies and urethral pressure profile (UPP)
- 51729: Complex CMG with voiding pressure studies, UPP, and EMG
- 51741: Complex uroflowmetry (electronic equipment)

To obtain this data:
1. Go to CMS Medicare Provider Utilization and Payment Data (Physician and Other Practitioners): https://data.cms.gov/provider-summary-by-type-of-service/medicare-physician-other-practitioners/medicare-physician-other-practitioners-by-provider-and-service
2. Filter by HCPCS codes: 51726, 51727, 51728, 51729, 51741
3. Download the filtered dataset as "Medicare Urodynamics Data.xlsx"

```{r}
x <- fread("Medicare_Urodynamics_Data.csv") %>% data.table
```

### Clean up

#### First round

```{r}
#Clean column names
x <- clean_names(x)
colnames(x) <- colnames(x) %>% str_remove_all("rndrng_|prvdr_")

#Create a column consisting of city and state
x[, city_state_zip := paste(city, state_abrvtn, zip5, sep = ", ")]

x[, full_original_st := fifelse(
  is.na(st2), st1, paste(st1, st2, sep = ", ")
)]

#Clean up the location variable
x[, st1 := st1 %>%
    str_replace_all(c("\\bSq\\b"  = "Square",
                      "\\bAve\\b"  = "Avenue",
                      "\\bPlz\\b"  = "Plaza",
                      "\\bRd\\b"  = "Road",
                      "\\bPkwy\\b"  = "Parkway",
                      "\\bBlvd\\b"  = "Boulevard",
                      "\\bBl\\b"  = "Boulevard",
                      "\\bSte\\b"  = "Street",
                      "\\bSt\\b"  = "Street",
                      "\\bFwy\\b"  = "Freeway",
                      "\\bDr\\b"  = "Drive",
                      "\\bCir\\b"  = "Circle",
                      "\\bBldg\\b" = "Building",
                      "\\bLn\\b" = "Lane",
                      "\\bHls\\b" = "Hills",
                      "\\bTrl\\b" = "Trail",
                      "\\bCtr\\b" = "Center",
                      "\\bHwy\\b" = "Highway",
                      "\\bMem\\b" = "Memorial",
                      "\\." = "",
                      "," = "",
                      "-" = "",
                      "#" = "",
                      " [0-9+]+$" = "",
                      "(?i) [a-z+][0-9+]+$" = "",
                      " .$" = "",
                      "(?<=[0-9])[a-z]" = "")) %>%
    str_remove_all("Square|Avenue|Plaza|Road|Boulevard|Street|Drive|Parkway|Freeway|Suite|Circle|Building|Lane|Hills|Trail|Highway|\\b..\\b") %>%
    str_squish]
```

#### Coalesce by location

```{r}
#Create a location variable consisting of street address, city, and state
x[, address := paste(st1, city, state_abrvtn, zip5, sep = ", ")]

#Coalesce by location
locs <- x[, .(tot_benes = sum(tot_benes),
      tot_srvcs = sum(tot_srvcs),
      avg_mdcr_pymt_amt = mean(avg_mdcr_pymt_amt),
      city = city %>% first,
      st = full_original_st %>% first,
      state = state_abrvtn %>% first,
      zip5 = zip5 %>% first),
  by = address]

#Recreate address using original full street name
locs[, address := paste(st, city, state, zip5, sep = ", ")]
```

#### Second round

After initial geocoding, manually inspect for duplicate locations with slightly
different addresses and consolidate them here. Since urodynamics data will have
different providers than cardiac PET, this section needs to be populated after
reviewing the actual data.

```{r}
# NOTE: After downloading and inspecting the urodynamics Medicare data,
# add manual address corrections here to consolidate duplicate locations.
# Example format:
# locs[, address := address %>%
#        str_replace_all(c(
#          "old address 1" = "corrected address 1",
#          "old address 2" = "corrected address 2"
#        ))]
```

#### Recoalesce

```{r}
#Coalesce by location
locs <- locs[, .(tot_benes = sum(tot_benes),
      tot_srvcs = sum(tot_srvcs),
      avg_mdcr_pymt_amt = mean(avg_mdcr_pymt_amt),
      city = city %>% first,
      st = st %>% first,
      state = state %>% first,
      zip5 = zip5 %>% first),
  by = address]
```

### Geocode

```{r}
#Get geocodes (longitude/latitude)
geo_codes_arc <- locs %>%
  geocode(address = address,
          method = "arcgis",
          verbose = TRUE)

#Coalesce by state
states_n <- locs[, .(n = .N), by = state]

#Add states with 0 urodynamics centers
states_n <- states_n[, state.abb[state.abb %nin% state]] %>%
  add_row(states_n, state = ., n = 0)
```

## Plot

### Box plot

```{r}
f1 <- ggplot(data = locs,
       aes(y = tot_benes)) +
  geom_beeswarm(data = locs,
                alpha = 0.4,
                size = 6,
                cex = 4,
                color = "lightblue3",
                aes(y = tot_benes, x= 0)) +

  geom_boxplot(outlier.shape = NA,
               fill =  NA,
               color = "maroon",
               linewidth = 2) +
  scale_y_continuous(trans = "log",
                     name = "Total number of beneficiaries (patients)",
                     breaks = c(10, 20, 50,
                                100, 200, 500,
                                1000, 2000, 5000,
                                10000)) +
  scale_x_continuous(name = NULL, labels = NULL, breaks = NULL) +
  theme_pubclean() +
  theme(text = element_text(size = 23, face = "bold"),
        plot.title=element_text(face = "bold",hjust = 0.0, size = 20),
        plot.subtitle = element_text(face = "bold", size = 13, hjust = 0.0, color = "grey45"),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_text(size = 20, face = "bold"),
        axis.title.x = element_text(size = 25, face = "bold"),
        axis.title.y = element_text(size = 25, face = "bold"),
        axis.line = element_line(colour = "black", linewidth = 1.2),
        legend.position = "right",
        legend.key.size = unit(1.25, "cm"))

ggsave(f1, filename = "Boxplot.png",
       dpi = 600,
       height = 10, width = 10)
```

### State-level N

```{r}
#Transform latitude/longitude into acceptable format by usmap
points_df <- usmap_transform(data = geo_codes_arc,
                input_names = c("long", "lat"))

#Plot!
f2 <- plot_usmap(data = states_n,
                 values = "n") +
  scale_fill_gradient_tableau(name = "Number of urodynamics testing centers") +
  geom_point(data = points_df,
             aes(x = x, y = y),
             col = "maroon",
             alpha = 0.4,
             size = 5) +
    ggtitle("State-level availability of urodynamics testing centers") +
  theme(
    legend.text = element_text(size = 18, face = "bold"),
    legend.title = element_text(size = 16, face = "bold"),
    legend.background = element_blank(),
    legend.key = element_blank(),
    legend.position = c(0.6, 0.001),
    plot.title=element_text(face = "bold", hjust = 0.5, size = 25),
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.title = element_blank()) +
  guides(fill = guide_colourbar(direction = "horizontal", title.position = "top",
                                barwidth = 20, barheight = 3,
                                ticks.colour = "black", ticks.linewidth = 0,
                                override.aes = list(size = 12))
  )

ggsave(filename = "N per state.png",
       bg = "white",
       dpi = 600,
       width = 16,
       height = 9)
```

### State-level density (Figure 1A)

```{r}
#Bind full state names
states_n <- states_n %>%
  left_join(data.frame(state_full = state.name,
                       state_abb = state.abb),
                   by = c("state" = "state_abb"))

#Get US pop size
state_pop <- get_decennial(geography = "state",
                           year = 2020,
                           variables = "P1_001N")

#Join
states_n <- states_n %>%
  left_join(state_pop %>%
              select(NAME, value) %>%
              rename(pop = value),
                   by = c("state_full" = "NAME"))

#Transform latitude/longitude into acceptable format by usmap
points_df <- usmap_transform(data = geo_codes_arc,
                input_names = c("long", "lat"))

#Add DC data
states_n[state == "DC",
         `:=`(state_full = "District of Columbia",
              pop = 689545)]

#Add urodynamics density column
states_n[, uro_density := 1000000*(n/pop)]

#Arrange state by density
states_n[, state := factor(state, states_n %>% arrange(-uro_density) %>% pull(state))]

#Plot!
f1a <- plot_usmap(data = states_n,
                 values = "uro_density") +
    geom_point(data = points_df,
             aes(x = x, y = y),
             col = "grey20",
             alpha = 0.4,
             size = 3) +
  scale_fill_distiller(palette = "RdYlBu", direction = +1,
                       name = "Number of urodynamics testing centers per million") +
    ggtitle(waiver()) +
  theme(
    legend.text = element_text(size = 18, face = "bold"),
    legend.title = element_text(size = 16, face = "bold"),
    legend.background = element_blank(),
    legend.key = element_blank(),
    legend.position = c(0.6, 0.001),
    plot.title=element_text(face = "bold", hjust = 0.5, size = 25),
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.title = element_blank()) +
  guides(fill = guide_colourbar(direction = "horizontal", title.position = "top",
                                barwidth = 20, barheight = 3,
                                ticks.colour = "black", ticks.linewidth = 0,
                                override.aes = list(size = 12))
  ) +
  force_panelsizes(rows = unit(9, "in"),
                   cols = unit(16, "in"))

ggsave(filename = "Figure 1A.png",
       bg = "white",
       dpi = 1000,
       width = 16,
       height = 9)
```

### Intersection of points with rural/urban areas

```{r}
#Read rural-urban classification data
rudata <- sf::read_sf("RU Data/tl_2021_us_cbsa.shp")

#Rename long/lat columns
rudata <- rudata %>%
  rename(lon = INTPTLON, lat = INTPTLAT) %>%
  mutate(across(.cols = c("lon", "lat"), ~ as.numeric(.))) %>%
  clean_names %>%
  filter(lsad == "M1")

#Convert the object returned by tidygeocode into an sf object
geo_uro_points <- geo_codes_arc %>%
  st_as_sf(coords = c("long", "lat"))

#Create matching CRS
st_crs(geo_uro_points) <- "WGS84"
geo_uro_points <- st_transform(geo_uro_points, crs = "NAD83")

#Get N of intersections
ints <- st_intersects(rudata, geo_uro_points)

#Sum them up
metro <- lapply(1:length(ints),
       function(i) length(ints[[i]]) > 0) %>%
  unlist %>%
  factor %>%
  summary

#Convert to data.table
rudata <- rudata %>% data.table

#Confirm in another way
for(i in 1:rudata[,.N]) {

  any_present <- which(locs$city %in% rudata[i, name %>%
           word(1, sep = ",") %>%
           str_split("-") %>%
           unlist])

  if(length(any_present > 0)) {
  same_state <- locs$state[any_present] %in% rudata[i, name %>%
           word(2, sep = ",") %>%
           str_split("-") %>%
           unlist %>% str_trim]
  }

  if(length(any_present > 0) & any(same_state) ) {
    rudata[i, present := TRUE]
  } else {
    rudata[i, present := FALSE]
  }

}

#Compare estimates
rudata[, present %>% summary]
metro
```

### Census block-based map

#### Prepare data

```{r}
#Import census tract data
gdata <- sf::st_read(dsn = "tlgdb_2021_a_us_substategeo.gdb",
                  layer = "Census_Tract")

#Rename long/lat columns
gdata <- gdata %>%
  rename(lon = INTPTLON, lat = INTPTLAT) %>%
  mutate(across(.cols = c("lon", "lat"), ~ as.numeric(.)))

#Create a matrix containing long/lat of census tracts
ct_matrix <- matrix(ncol = 2, nrow = nrow(gdata),
       c(gdata$lon,
         gdata$lat)
         )

#Create a matrix containing long/lat of urodynamics centers
uro_matrix <- matrix(ncol = 2, nrow = nrow(points_df),
       c(points_df$long,
         points_df$lat)
         )

#Calculate distance using the Haversine function
distances <- distm(x = uro_matrix, y = ct_matrix,
                   fun = geosphere::distHaversine)

#Create an empty min_distances vessel
min_distances <- as.numeric()

#Fill up the vessel with the distance between each census tract and the nearest urodynamics center
for(i in 1:nrow(ct_matrix)) {
 min_distances <- c(min_distances, distances[, i] %>% min)
}

#Convert to km
min_distances <- (min_distances/1000)

#Remove redundant object
rm(distances)

#Attach
gdata$dist <- min_distances

#Import census tract file
ct_data <- fread("US Census Tract 2020 Data.csv",
                 header = TRUE,
                 select = c("GEO_ID", "NAME", "DP1_0001C"))

#Remove the first 2 rows (non-census tract data)
ct_data <- ct_data[!c(1, 2),]

#Rename columns
ct_data <- ct_data %>%
  rename(any_of(.q(pop = DP1_0001C,
                   GEOID = GEO_ID)))

#Extract state names
ct_data[, state := fifelse(NAME %>% str_detect(";"),
                           word(NAME, -1, sep = "; "),
                           word(NAME, -1, sep = ", "))]

#Format GEOID similar to how it is in gdata
ct_data[, GEOID := GEOID %>% str_remove("1400000US")]

#Merge together
gdata <- merge(gdata,
               ct_data,
               by = "GEOID")

#Remove ct_data
rm(ct_data)

#Make population numeric
gdata <- gdata %>%
  mutate(pop = pop %>% as.numeric)
```

#### Plot map (Figure 1B)

##### Minus PR/HW/AK

```{r}
f1b <- ggplot(data = gdata %>%
               filter(state %nin% c("Hawaii", "Alaska", "Puerto Rico")),
             aes(fill = cut(dist,
                            c(0, 5, 10, 25, 50, 100, 200, 500, 1000, 100000)),
                  color = after_scale(fill))) +
  geom_sf(color = NA, linewidth = 0) +
  scale_fill_brewer(palette = "RdYlBu", direction = -1,
                    name = "Distance (kilometres)",
                    labels = c("Less than 5",
                               "5 to 10",
                               "10 to 25",
                               "25 to 50",
                               "50 to 100",
                               "100 to 200",
                               "200 to 500",
                               "500 or greater")) +
    ggtitle(waiver()) +
  theme(
    legend.text = element_text(size = 18, face = "bold"),
    legend.title = element_text(size = 16, face = "bold"),
    legend.background = element_blank(),
    legend.key = element_blank(),
    legend.position = c(0.9, 0.25),
    plot.title=element_text(face = "bold", hjust = 0.5, size = 25),
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.background = element_blank(),
    axis.title = element_blank())
```

##### Puerto Rico

```{r}
pr <- ggplot(data = gdata %>%
               filter(state %in% c("Puerto Rico")),
             aes(fill = cut(dist,
                            c(0, 5, 10, 25, 50, 100, 200, 500, 1000, 100000)),
                  color = after_scale(fill))) +
  geom_sf(color = NA, linewidth = 0) +
  scale_fill_brewer(palette = "RdYlBu", direction = -1,
                    name = "Distance (kilometres)",
                    labels = c("Less than 5",
                               "5 to 10",
                               "10 to 25",
                               "25 to 50",
                               "50 to 100",
                               "100 to 200",
                               "200 to 500",
                               "500 or greater")) +
    theme(
    legend.text = element_text(size = 18, face = "bold"),
    legend.title = element_text(size = 16, face = "bold"),
    legend.background = element_blank(),
    legend.key = element_blank(),
    legend.position = "none",
    plot.title=element_text(face = "bold", hjust = 0.5, size = 25),
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.background = element_blank(),
    plot.background = element_blank(),
    axis.title = element_blank()) +
  coord_sf(xlim = c(-68.5, -64.5),
           ylim = c(17.8, 18.8)) +
  force_panelsizes(rows = unit(0.5, "in"),
                   cols = unit(0.8, "in"))
```

##### Hawaii

```{r}
hw <- ggplot(data = gdata %>%
               filter(state %in% c("Hawaii")),
             aes(fill = cut(dist,
                            c(0, 5, 10, 25, 50, 100, 200, 500, 1000, 100000)),
                  color = after_scale(fill))) +
  geom_sf(color = NA, linewidth = 0) +
  scale_fill_brewer(palette = "RdYlBu", direction = -1,
                    name = "Distance (kilometres)",
                    labels = c("Less than 5",
                               "5 to 10",
                               "10 to 25",
                               "25 to 50",
                               "50 to 100",
                               "100 to 200",
                               "200 to 500",
                               "500 or greater")) +
    theme(
    legend.text = element_text(size = 18, face = "bold"),
    legend.title = element_text(size = 16, face = "bold"),
    legend.background = element_blank(),
    legend.key = element_blank(),
    legend.position = "none",
    plot.title=element_text(face = "bold", hjust = 0.5, size = 25),
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.background = element_blank(),
    plot.background = element_blank(),
    axis.title = element_blank()) +
  coord_sf(xlim = c(-180, -150),
           ylim = c(18, 30)) +
  force_panelsizes(rows = unit(2, "in"),
                   cols = unit(3.2, "in"))
```

##### AK

```{r}
ak <- ggplot(data = gdata %>%
               filter(state %in% c("Alaska")),
             aes(fill = cut(dist,
                            c(0, 5, 10, 25, 50, 100, 200, 500, 1000, 100000)),
                  color = after_scale(fill))) +
  geom_sf(color = NA, linewidth = 0) +
  scale_fill_brewer(palette = "RdYlBu", direction = -1,
                    name = "Distance (kilometres)",
                    labels = c("Less than 5",
                               "5 to 10",
                               "10 to 25",
                               "25 to 50",
                               "50 to 100",
                               "100 to 200",
                               "200 to 500",
                               "500 or greater")) +
    theme(
    legend.text = element_text(size = 18, face = "bold"),
    legend.title = element_text(size = 16, face = "bold"),
    legend.background = element_blank(),
    legend.key = element_blank(),
    legend.position = "none",
    plot.title=element_text(face = "bold", hjust = 0.5, size = 25),
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.background = element_blank(),
    plot.background = element_blank(),
    axis.title = element_blank()) +
  coord_sf(xlim = c(-180, -130),
           ylim = c (51, 72)) +
  force_panelsizes(rows = unit(2, "in"),
                   cols = unit(3.2, "in"))
```

##### Merge

```{r}
#Merge with Alaska
f1b <- f1b + inset_element(ak,
                            left = 0, right = 0.3, top = 0.3, bottom = 0)

#Merge with Hawaii
f1b <- f1b + inset_element(hw,
                            left = 0.2, right = 0.4, top = 0.3, bottom = 0)

#Merge with Puerto Rico
f1b <- f1b + inset_element(pr,
                            left = 0.75, right = 0.85, top = 0.1, bottom = 0)

#Save plot!
ggsave(filename = "Figure 1B.png",
       dpi = 1000,
       width = 16,
       height = 9)
```

#### Get relevant numbers

```{r}
##N
#Total number of unique urodynamics testing centers
locs[, uniqueN(address)]

#By State
states_n %>% arrange(-n)

#States with no urodynamics
states_n[n == 0]

##Density
states_n %>%
  arrange(-uro_density) %>%
  mutate(uro_density = round(uro_density, 1))

##Median distance
#Remove geom
gdata <- gdata %>%
  select(pop, state, dist) %>%
  data.table

#Calculated weighted median distance
gdata[, wtd.quantile(dist, pop, probs = c(0.25, 0.5, 0.75))
      %>% round(1)]

##N and % of US citizens more than 100km from a urodynamics testing center
round(gdata[dist > 100, pop %>% sum]/1000000, 1) #N (in millions)
round(100*gdata[dist > 100, pop %>% sum]/gdata[, pop %>% sum], 1) #Percentage

#Nearest by state
gdata[,
      .(mdist = wtd.quantile(dist, pop, probs = 0.5) %>% round(0)),
      by = state] %>% arrange(-mdist)

#N of metropolitan statistical areas with a urodynamics testing center
metro
round(100*metro[1]/sum(metro[1:2]), 1)
```
